---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('bppg')`"
vignette: >
    %\VignetteIndexEntry{1. Introduction to BPPG}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("BiocStyle")
library("RefManageR")
library("bppg")

## Write bibliography information
file <- system.file("extdata", "ref.bib", package = "bppg")
bib <- ReadBib(file, check = FALSE)

bib <- as.BibEntry(c(citation(), citation("BiocStyle")[1], citation("knitr")[1],
    citation("RefManageR")[1], citation("rmarkdown")[1], 
    citation("sessioninfo")[1], citation("testthat")[1],
    citation("ggplot2")[1], citation("Rsolnp")[1],
    citation("seqinr")[1], citation("batchtools")[1], citation("bppg")[1], 
    citation("bppg")[2])) + 
    bib

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```



# Introduction

<!-- This is kind of long and should be  moved into the matching section
ALso, add bullet points for steps/what it does and what is good about it -->

Bottom-up mass spectrometry (MS) is an effective tool for analyzing 
high-throughput proteomics. This method allows the proteome of cells, tissues 
and organisms to be investigated in depth. It allows diseases such as 
Parkinson's and the effects of stress on cells and organisms to be studied at a 
proteomic level. In bottom-up MS, however, proteins are digested into smaller 
peptides, which are then measured. Consequently, the measured intensities 
correspond to peptides rather than proteins. While this makes differential 
analysis at the peptide level possible, it lacks biological interpretability. 
Peptide intensities can instead be aggregated to protein intensities, but this 
is not straightforward due to shared peptides. 

Our package, `bppg` (**b**ipartite **p**eptide **p**rotein **g**raphs), is 
designed to utilize all peptide-level information for protein quantification. 
This is achieved by modelling the relationship between the quantified peptides 
and their proteins of origin using bipartite graphs. These graphs form the basis
of an equation system to which an optimization method is applied  to provide
the best explanation of  the measured ratios using the optimized protein ratios.

The functions and steps of the workflow, from peptide quantities to relative 
protein quantifications, include:

- Reading data of peptide quantities from different software
- Normalization of peptide quantities
- Aggregation of replicates per sample group
- Calculation of peptide ratios per group comparison 
- Build bipartite graphs per comparison
- Optimization step to calculate protein ratios 

In this vignette we included a quick start guide to use `bppg`.

## Background - bppg Theory

### Peptides

There are two categories of peptides. First, **unique** peptides, which 
originate from one protein or protein group (proteins that cannot be 
distinguished based on the measured peptides). Second, **shared** peptides, 
which can derive from more than one protein/protein group. **Unique** peptides
are strong evidence for the presence of a protein and 
**shared** peptides provide ambiguous information. 
`r Citep(bib[["schork2024improvement"]])`

### Bipartite Graphs

In a bipartite graph, there are two types of nodes (in this case, proteins and
peptides) and edges only exist between different types, never within one type of
group. A protein node is connected to all their possible 
peptide products. **Unique** peptides therefore have only one edge, while 
**shared** peptides have several. This graph can be based on only *in silico* 
information, creating a theoretical graph.
`r Citep(bib[["schork2024improvement"]])`

This example shows two proteins, which share one peptide and also each have an
unique peptide.
```{r example_plot, echo = FALSE, message = FALSE, out.width="70%", fig.asp=0.9}
file <- system.file("extdata", "quantGraphs_collpept.rds", package = "bppg")
G_coll <- readRDS(file)[[1]][[3]]
plotBipartiteGraph(G_coll)
```


# Installation of `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }

BiocManager::install("bppg")

# or 
pak::pak("bppg")
```


# Asking for help

If you need help with `bppg`, encounter a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r cite}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of `bppg` is protein inference and relative protein quantification
for bottom-up mass spectrometry. This method is only suitable for relative
analysis. Therefore, at least two samples are required to calculate the 
protein quantities. However, at least 3 replicates per group are recommended. 

**Required inputs**:

- Peptide intensities
    - Output from MaxQuant or FragPipe (more options are in development)
    - The read function creates a custom SummarizedExperiment with sample 
    metadata in `colData` and peptide information in `rowData`
- The FASTA file used for peptide identification


## Data

### Example Data
A test dataset and the matching FASTA file are provided within the package. It 
is a subset of a benchmark dataset which was created to determine the 
correctness 
of protein quantification software `r Citep(bib[["ramus_spiked_2016"]])`. 
The original dataset includes: 

- Yeast background, 48 UPS spike-ins
- 9 different spike-in concentrations
- 3 replicates each = 27 samples total

The example data used in this vignette includes peptides from 9 yeast proteins.

### Import
The peptide quantification was performed with MaxQuant (version 2.7.3.0)
and resulted in a `peptides.txt` file. The
corresponding function to extract the columns of interest is
`readMqPeptideTable`, this includes the peptide sequences and intensity values.

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides_filtered.txt", package = "bppg")
D <- readMqPeptideTable(path = file)

```

Here is an example of the structure in our SummarizedExperiment dataset:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D  # SummarizedExperiment overview
head(colData(D))  # sample/replicate metadata
assays(D)$intensities[1:5, 1:3]  # peptide intensity matrix 
head(rowData(D))  # peptide metadata (sequence, protein, etc.)
```


## FASTA file

To build the bipartite graph, proteins need to be digested *in silico* and 
peptides matched with their protein of origin. 
For this we have a reduced FASTA file.

```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta",
    package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)
names(fasta)
```

## Build Quantification Graph

Now the theoretical information can be combined with quantification data into
a bipartite peptide protein graph. 
To simplify this, we created a wrapper function, which includes digestion 
and data
preprocessing (normalization, replicate aggregation and ratio calculation).
To generate the quantification graphs, the protein nodes are collapsed
(`collProtNodes = TRUE`), this creates protein groups of proteins which
contain the same measured peptides.


```{r build_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- bppg::generateGraphsFromQuantData(D = D,
    fasta = fasta,
    collProtNodes = TRUE)

```

The graph created for each
comparison is not fully interconnected but many independent graphs which are
considered separately in the downstream analysis.
Therefore, `graphs` is a nested list, first on the level of comparisons and 
then the multitude of independent graphs for that comparison. 

`graphs[[1]][[3]]` shows one graph of the first comparison 50 amol vs 125
amol. The `igraph` object shows the amount of nodes and edges, the attributes
and the first edges. This example is from the RIR2 protein which does not have 
an overlap with another protein. The peptide nodes show the respective 
log2-transformed peptide ratios.

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[3]]
```

```{r graphs_plot, message = FALSE, out.width="70%", fig.asp=0.9}
plotBipartiteGraph(graphs[[1]][[3]], node_labels_proteins = "accessions",
    node_labels_peptides = "pep_logRatios")
```


## Optimize Relative Quantification

The optimization is performed independently on each graph. The bipartite graph 
forms the basis of an equation system used to calculate protein level ratios 
($R_i$). 
The graph combines information about the peptide ratios ($r_j$) and 
the proteins from which they can originate.
However, this is not enough to calculate them directly, instead, a weight 
($C_i$) is needed for each protein. 
The optimization algorithm `Rsolnp` `r Citep(bib[["galanos2025rsolnp"]])` is 
applied to the resulting equation system.

As an example we will use the following graph containing the proteins RL13A and 
RL13B.

```{r plot_opti, message = FALSE, out.width="70%", fig.asp=0.9}
plotBipartiteGraph(graphs[[1]][[1]], node_labels_proteins = "accessions",
    node_labels_peptides = "pep_logRatios")
```

The optimization process involves two steps.
First, different combinations of protein weights are tested. 
In the second step, the best fitting (smallest error) $C_i$ combination is 
extracted.

```{r optimization,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
res <- iterateOverCi(graphs[[1]][[1]]) 
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[1]], res)
```

Some of the results are not single values but a 
range of equally optimal solutions. This occurs if several combinations return 
the same error and the result is ambiguous. 
This can occur if peptide ratios are contradicting or the equation system is
underdetermined. 

The result is in a `SummarizedExperiment` object, and the table can be accessed 
via the assays function.

```{r res_optimization, eval = TRUE, include = TRUE, message = FALSE}
head(assays(protQuant)$results)
```
In this example, the `RiLog_min` and `RiLog_max`, as well as the `Ci_min` and 
`Ci_max` can be seen. This is how the range results are encoded. These values 
represent the boundaries of the log ratio and range for the given error.
The `error_min` shows how well the calculated protein ratios fit the measured
peptide ratios.

Considering the example graph, the reasons for the error value and
ambiguity of the results are evident. The log(ratio) of the shared peptides
ranges from -0.6 to 2.63, with a mean of around 0.55. The resulting variability 
explains the high margin of error. The log(ratio) of  RL13B must be close to its
unique peptide at 0.61. The size of the effect of RL13B on the shared peptides 
depends on the value of $C_i$. Therefore, the log(ratio) of RL13A must adapt to 
this, creating the range of possibilities.



## Combine and Export
Results from the same comparison can be combined by using `rbind()` on the quant
results. However, running the optimization for all the sample graphs takes too 
long for this vignette. 

The results from all the graphs can be combined into a single result. Each 
comparison is treated as one assay in the final `SummarizedExperiment`.

Please note that for simplicity, the comparisons in this example file are
called "1_2", "1_3" etc instead of using the spike-in concetration levels.

```{r combine,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
file <- system.file("extdata", "resultsList.rds", package = "bppg")
resultList <- readRDS(file)
SE <- combineComparisons(resultList)
SE
```

It can also be exported to an Excel spreadsheet, with the assays on different 
sheets.
```{r export,  echo = TRUE, eval = FALSE, include = TRUE, message = FALSE}
exportSE(SE, file.path("path_to_folder","results.xlsx"))
```

### Parallelization

Although the single example was quick, processing a complete dataset can take 
hours 
due to the large number of graphs involved. However, optimization is applied to
each graph individually, enabling easy parallelization using tools such as
`batchtools` `r Citep(bib[["lang2017batchtools"]])`, which supports HPC job 
scheduling. 
You can see an example of this in the Nextflow workflow for this package:
[here](https://github.com/mpc-bioinformatics/bppg-nf/blob/devel).

# Further information

- More detailed information on the parameters and settings can be found 
in the documentation of the respective functions 
`r Citep(bib[["schork2026bppg"]])`
- Information on the detailed workflow, *in silico* analysis and customizing the
graphics are planned for future vignettes.

# Comparison with other related packages

The amount of R packages for mass spectrometry is continuously increasing. Many
are considering raw spectrum processing or downstream statistical analysis. 
However, the amount of packages that include methods for protein inference and
quantification are limited and mostly confined to simple peptide aggregation 
approaches. 

With `bppg` we provide a more transparent quantification approach that also 
handles proteins without unique peptides.
There are some packages with functionalities for proteomics data that cover 
other parts of the whole workflow. This could include peptide identification 
approaches or further statistical analysis.

- `MSstats` is a package for statistical analysis of proteomics data. Its 
extension `MSstats+` quantifies proteins from peptide abundances by aggregating
them with a weighted sum based on the peak quality.
- `MSnbase` processes proteomics data from raw to quantification data. For the 
protein quantification a median and weighted average function are implemented.
- `msqrob2` analyses differential abundance with a hurdle model. It can be run 
with peptide quantities, which are later aggregated with `iq::maxLFQ`.
- `mzR` provides parsing for different standard file formats used for mass
spectrometry data.
- `DEqMS` is implemented on top of limma and expects protein variance dependent 
on quantified peptides and PSMs.
- `QFeatures` implements the QFeature object consisting of a 
SummarizedExperiment and provides functionality for processing and adding meta 
data.

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```


# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["ole2025biocstyle"]])` with `r CRANpkg("knitr")` 
`r Citep(bib[["xie2025knitr"]])` and `r CRANpkg("rmarkdown")` 
`r Citep(bib[["allaire2026rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` 
`r Citep(bib[["mclean2017refmanager"]])`.


```{r Biblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography

PrintBibliography(bib, .opts = list(hyperlink = "to.bib", style = "html"))
```
