---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{1. Introduction to BPPG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    ggplot2 = citation("ggplot2")[1],
    Rsolnp = citation("Rsolnp")[1],
    seqinr = citation("seqinr")[1],
    batchtools = citation("batchtools")[1]
)

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```

# Introduction

<!-- This is kind of long and should be  moved into the matching section
ALso, add bullet points for steps/what it does and what si good about it -->

Bottom-up mass spectrometry (MS) is an effective tool for analysing
high-throughput proteomics. This method enables the comprehensive investigation
of the proteome of cells, tissues, and organisms. This allows diseases such as 
Parkinson's and the effects of stress to be studied at a proteomic level. 
However, in bottom-up MS, proteins are digested into smaller peptides, which 
are then measured. Consequently, the resulting abundances correspond to 
peptides, rather than proteins. Differential peptide analysis can be performed 
between two sample groups.

Our package **b**ipartite **p**eptide **p**rotein **g**raphs (`bppg`) aims to
preserve the peptide ratios and built on them for the protein quatification. 
This is done by using bipartite graphs to model the relationship between the 
quantified peptides and their proteins. Which builts the basis for an equation
system which is then optimized to best explain  the meassured ratios with 
protein ratios. The optimization allows for inconsistencies between peptides of 
the same set of proteins. Due to the nature of MS and biological samples, the 
peptide intensities are highly variable and can lead to contradictions. 

The functions and steps include:

- Reading MaxQuant data of peptide quantities
- Normalization of peptide quantities
- Aggregation of replicates per sample group
- Calculation of peptide ratios per group comparison
- Built bipartite graph per comparison
- Optimize for protein ratio/fold change

More functions are described in additional vignettes:

- Built insilico bipartite graph (WiP, 2. Analysing *inSilico* graphs)
- Plot bipartite graphs (WiP, 3. Visualize `bppg`)

# Installation of  `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
pak::pak("mpc-bioinformatics/bppg")
```


# Asking for help

If you need help with `bppg`, encountered a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r "citation"}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of `bppg` is protein inference and relative protein quantification
for bottum-up mass spectrometry. This methods is only suitable for relative
analysis. Therefore, at least two samples are requiered to calculate the 
protein quantities. Ideally, with several replicates per sample. 

`bppg` works with peptide intensities and currently supports the data format
from
MaxQuant. Otherwise, the dataset can be formatted independtly into
a SummarizedExperiment object with sample/replicate informations in the
columns/colData and peptide information in the rows/rowData.
Additionally, `bppg` requiers the FASTA file which was used for the peptide
identification.

The second application of `bppg` is the modeling of peptide and protein 
relationships using bipartite graphs. This is part of the quantification
workflow, but can be used independtly for an insilico analysis. It is 
described in a second vignette.

## Data

### Example Data
A test dataset and the matching FASTA file is provided with the package. It is
a subset of a gold standarad which was created to determine the correctness of
protein quantification software.
The original dataset includes: 

- Yeast background, 48 UPS spike in
- 9 different spike in concentrations
- 3 replicates each = 27 samples total

The example data for this workflow only includes peptides from a subset of 
yeast proteins.

### Import
For this dataset the peptide quantificationwas was performed with MaxQuant. The
corresponding function to extract the columns of interest is
`readMqPeptideTable`.
Example of loading a `peptides.txt` file:

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides_filtered.txt", package = "bppg")
D <- readMqPeptideTable(path = file, LFQ = TRUE, remove_contaminants = FALSE)

```

The SummarizedExperiment dataset should look as followed:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D
head(colData(D))
assays(D)$intensities[1:5, 1:3]
head(rowData(D))
```

## Data Preprocessing 

### Normalization

Normalization (`normalizePeptideIntensities`) has to be performed to avoid batch
effects. The default method is 
`loess`, other methods include: `median`, `quantile` or `lts` normalization. 
If the data is already normalized, there is an option for `nonorm` (no 
normalization). Nevertheless, the function has to be called to allow for 
structural change in the SummarizedExperiment object.

```{r normalize,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D_norm <- normalizePeptideIntensities(D, method = "loess")

```


### Aggregation
To be able to calculate realtive protein quantities, the peptides have to be 
aggregated. This avoids the need to build
graphs for each combination of replicates. The replicates need to be assigned 
their sample group and then they are aggregated, e.g. by mean. If the group is 
not a 
subset of the sample name it is recommended to write a factor vector with the
order corresponding to the samples.

The `missing.limit` is used to determine in how many replicates needa peptide 
needs to be quantified to be counted as not missing. If it 
is set to 0.0 it allows no missing in the group, while 1.0 allows all 
missing. The default is 0.0.

```{r aggregation,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
group <- factor(limma::strsplit2(colnames(D_norm), split = "_")[, 1])
levels(group)
D_aggr <- aggregateReplicates(D_norm, method = "mean", missing.limit = 0.4,
    group = group, seq_col = "Sequence")

```

### Ratio

Next, the ratio or fold change between the aggregated intensities can be
calculated. The comparison on peptide level removes confounding factors, e.g.
ionization efficiancy. The function `calculatePeptideRatios` calculates all
pair-wise group comparisons combinations. These peptide ratios will be the
basis of the bipartite graph.

```{r ratio,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
groups  <- levels(group)
peptide_ratios <- calculatePeptideRatios(D = D_aggr, group_levels = groups)
colData(peptide_ratios)
```


## FASTA

To built the bipartite graph, peptides need to be *insilico* digested and 
matched 
with their protein of origin. This is done by the function `digestFASTA`, which 
allows for the control of min and max amino acid length of a peptide and how
many miscleavages to consider. Additionally, with the fasta sequences a list of
species can be inputed (`protOrigin`), in the same order as the fasta entries.

```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta", package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)

edgelist <- digestFASTA(fasta)
head(edgelist)
```


## Built Graph

### bppg Theory
There are two categories of peptides. First, **unique** peptides, which 
originate from one protein or protein group. Second, **shared** peptides, 
which can derive from more than one protein/protein group. **Unique** peptides
are strong evidence for the abundance of a protein and while information from
**shared** peptides is more ambigious. 

Bipartite graphs have two types of nodes (proteins and peptides) and all edges
are between the two types. A protein node is connected to all their possible 
peptide products. **Unique** peptides therefore have only one edge, while 
**shared** peptides have several. This graph can be based on only *insilico* 
information, creating a theoretical graph. Alternatively, meassurements can be 
included, therefore removing all peptides without a valid ratio in the 
sample comparison. This removes proteins without any quantified peptides and
decreases the amount of proteins with definite evidence. Therby creating more
protein groups.

### Application
For the graph the quantification data and the FASTA edeglist are combined. 
The `generateQuantGraphs` function creates a comparison specific graphs, 
including only peptides with valid ratios from that comparison and their 
corresponding proteins.
The option to collapse protein nodes (`collProtNodes`) or collapse peptide nodes
(`collPeptNodes`) is given to reduce the complexity in the graph. Nodes are 
collapsed together if they have the same set of edges. This means collapsing 
protein nodes shows protein groups and is always recommended. 
`collPeptNodes = TRUE` is  not recommended for quantification purpose, since  
each peptide has its own quantification information. 

```{r built_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- generateQuantGraphs(exp_peptide_ratios = peptide_ratios,
    fasta_edgelist = edgelist, seq_column = "Sequence",
    collProtNodes = TRUE,
    collPeptNodes = FALSE)

```

`graphs` is a list of lists, first on the level of comparisons and then the 
multitude of independt graphs for that comparison. The created graph for each
comparison is not fully interconnected but many independet graphs which are
considered separatly in the downstream analysis. 

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[1]]
```



## Optimize Realtive Quantification

The quantified graph is the basis for an equation system to calculate protein 
level ratios. The known factors for this are the peptide ratios ($r_i$) and 
from which proteins they can originate. The unknwon factors are the protein 
ratios ($R_i$) and how much each protein contributed to the quantity of their
peptides compared to the other proteins ($C_i$). $C_i$ is only needed for one 
sample (A), since the other can be calculated with $R_i$. In general, the 
resulting equation is not solvable, since it has more unknown variables than 
known ones. 
Instead an optimization algorithm (`Rsolnp` `r Citep(bib[["Rsolnp"]])`) is 
applied to the system. The error
is calculated on the difference between the calculated $r_i$ and the actual 
ratio.

The optimization is divided into two steps and is performed per graph.

### Optimization
First, different $Ci$ percentages are fixed. The $Ci$ describes the percentage 
distribution of peptides per graph in sample A.
`iterateOverCi` considers a wide range of possible $Ci$ combination by fixing 
one value. The remaining $Ci$ are part of the optimization as are the protein 
ratios. The optimization aims to reduce the error between calculated peptide 
ratios, based on the optimized values, and the measured peptide ratios. With 
the following equation:
\[\underset{R_i, C_i, i=1,\dots,m}{\mathrm{argmin}} 
    \left( \sum_{j=1}^m \varepsilon_j^2  \right)\]

\[ \varepsilon_j = \log(r_j) - \log \left( \sum_{i = 1}^m \left[ 
    \frac{R_i \cdot C_i \cdot \delta_{ij}}{\sum_{k = 1}^m C_k 
        \cdot \delta_{kj}} \right] \right)\]

```{r optimierung,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
res <- iterateOverCi(graphs[[1]][[1]])
head(res)
```

### Extract Best Results

From the previous function the results are filtered for the combination of 
`Ci`s, with the lowest error. Some of the results are not absolute but span a 
range. This occurs if several combinations return the same error and therefore,
the result is ambigious. Nevertheless, that mirrows the information on the
peptide level.

In this example the `RiLog_min` and `RiLog_max`, as the `Ci_min` and `Ci_max `
can be seen.

```{r results,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[1]], res)
head(assays(protQuant)$results)
```


### Parallization

The optimization runs per graph. This enables easy parallization, with e.g.
`r Citep(bib[["batchtools"]])`. An example for this can be seen in our nextflow 
workflow 
for this package.

# Further information

- Information for insilico analysis and plotting can be found in further 
vignettes (WiP)

## Nextflow
We also provide a nextflow workflow 
[here](https://github.com/mpc-bioinformatics/bppg-nf/blob/devel). Which allows 
for the usage of a wider 
audience.

# Comparison with other related packages
what is this research?

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["BiocStyle"]])` with `r CRANpkg("knitr")` 
`r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` 
`r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r Biblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```