---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to BPPG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    ggplot2 = citation("ggplot2")[1],
    Rsolnp = citation("Rsolnp")[1],
    seqinr = citation("seqinr")[1]
)

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```

# Introduction

Bottom up mass spectrometry


# Installation of  `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
pak::pak("mpc-bioinformatics/bppg")
```


# Asking for help

If you need help with `bppg`, encountered a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r "citation"}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of bppg is protein inference and relative quantification for 
bottum-up mass spectrometry. This methods is only suitable for relative
analysis.
Therefore, at least two sample groups are requiered to calculate the protein 
quantities. Ideally, with several replicates per sample. 

bppg works with peptide intensities and currently supports the data format from
MaxQaunt and McFragger. Otherwise, the dataset can be formatted independtly into
a SummarizedExperiment object with sample/replicate informations in the
columns/colData andpeptide information in the rows/rowData.
Additionally, bppg requiers the FASTA file which was used for the  peptide
identification.

The second application of bppg is the modeling of peptide and protein 
relationships using bipartite graphs. This is part of the quantification
workflow, but can be used independtly for an insilico graph.

## Data

### Example Data
A test dataset and the matching FASTA file is provided with the package. 
The data is from ????

- yeast background, UPS spike in
- 9 different spike in concentrations
- 3 replicates each = 27 samples total

### Import
For this dataset the peptide quantificationwas performed with MaxQuant.
Example of loading a `peptides.txt` file:

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides_filtered.txt", package = "bppg")
D <- readMqPeptideTable(path = file, LFQ = TRUE, remove_contaminants = FALSE)

```

The SummarizedExperiment dataset should look as followed:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D
head(colData(D))
assays(D)$intensities[1:5, 1:3]
head(rowData(D))
```

## Data Preprocessing 

### Normalization

Normalization has to be performed to avoid batch effects. The default method is 
loess, other methods include: median, quantile or lts. 
If the data is already normalized, there is an option for nonorm (no 
normalization). Nevertheless, the function has to be called to allow for 
structural change in the SummarizedExperiment object.

```{r normalize,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D_norm <- normalizePeptideIntensities(D, method = "loess")


```


### Aggregation
bppg currently only compares the mean of two groups. To avoid calculating a 
large amount of graphs. The replicates need to be assigned their group and then 
there are aggregated, e.g. by mean.
A `missing.limit` can be set to decide for one group and one peptide, if the 
value is regarded as missing, since missing.limit replicates did not quantify 
this peptide. 0.0 allows no missing in the group, while 1.0 allows all missing.

```{r aggregation,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
group <- factor(limma::strsplit2(colnames(D_norm), split = "_")[, 1])
levels(group)
D_aggr <- aggregateReplicates(D_norm, method = "mean", missing.limit = 0.4,
    group = group, seq_col = "Sequence")

```

### Ratio

In the next step the ratios for each group comparison are calculated:

```{r ratio,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
groups  <- levels(group)
peptide_ratios <- calculatePeptideRatios(D = D_aggr, group_levels = groups)

```


## FASTA

To built the bipartite graph peptides need to be insilico digested and matched 
with their protein of origin. This is done by `digestFASTA`, which allows for 
the control of min and max amino acid length of a peptide and how many miscleavages 
to consider.
```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta", package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)

edgelist <- digestFASTA(fasta)

```


## Built Graph
For the graph the quantification data and the FASTA edeglist are combined. 
The `generateQuantGraphs` function creates a comparison specific edgelists, 
including only peptides with valid ratio from that comparison and their 
corresponding proteins.
The option to collapse protein nodes (`collProtNodes`) or collapse peptide nodes
(`collPeptNodes`) is given to reduce the complexity in the graph. Nodes are 
collapsed together if they have the same set of edges. This means collapsing 
protein nodes creates protein groups and therefore, is always recommended. 
`collPeptNodes = TRUE` is  not recommended for quantification purpose, since each 
peptide has its own quantification information. 

```{r built_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- generateQuantGraphs(exp_peptide_ratios = peptide_ratios,
    fasta_edgelist = edgelist, seq_column = "Sequence",
    collProtNodes = TRUE,
    collPeptNodes = FALSE)

```

`graphs` is a list of lists, first on the level of comparisons and then the 
multitude of independtly graphs for that comparison

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[1]]
```



## Optimize Realtive Quantification

The optimization is divided into two steps and is performed per graph.

### Optimization
First, different `Ci` percentages are fixed. The `Ci` describes the percentage 
distribution of peptides per graph in sample A.
`iterateOverCi` considers a wide range of possible `Ci` combination by fixing 
one value. The remaining `Ci` are part of the optimization as are the protein 
ratios. The optimization aims to reduce the error between calculated peptide 
ratios, based on the optimized values, and the measured peptide ratios.

```{r optimierung,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
res <- iterateOverCi(graphs[[1]][[1]])
head(res)
```

### Extract Best Results

From the previous results are filtered for the best combination of `Ci`s, with
the lowest error.

```{r results,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[1]], res)
head(assays(protQuant)$results)
```


### Parallization

The optimization runs per graph. This enables easy parallization, for with
example `batchtools`. an example for this can be seen in our nextflow workflow 
for this package.

# Further information

- information fo insilico analysis and plotting can be found in further 
vignettes

## Nextflow
We also provide a nextflow workflow here. which allows for usage of a wider 
audiance

# Comparison with other related packages
what is this research?

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

