---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('bppg')`"
vignette: >
    %\VignetteIndexEntry{1. Introduction to BPPG}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("BiocStyle")
library("RefManageR")
library("bppg")

## Write bibliography information
file <- system.file("extdata", "ref.bib", package = "bppg")
bib <- ReadBib(file, check = FALSE)

bib <- as.BibEntry(c(citation(), citation("BiocStyle")[1], citation("knitr")[1],
    citation("RefManageR")[1], citation("rmarkdown")[1], 
    citation("sessioninfo")[1], citation("testthat")[1],
    citation("ggplot2")[1], citation("Rsolnp")[1],
    citation("seqinr")[1], citation("batchtools")[1], citation("bppg")[1], 
    citation("bppg")[2])) + 
    bib

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```



# Introduction

<!-- This is kind of long and should be  moved into the matching section
ALso, add bullet points for steps/what it does and what is good about it -->

Bottom-up mass spectrometry (MS) is an effective tool for analysing 
high-throughput proteomics. This method allows the proteome of cells, tissues 
and organisms to be investigated in depth. It allows diseases such as 
Parkinson's and the effects of stress on cells and organisms to be studied at a 
proteomic level. In bottom-up MS, however, proteins are digested into smaller 
peptides, which are then measured. Consequently, the measured intensities 
correspond to peptides rather than proteins. While this makes differential 
analysis at the peptide level possible, it lacks biological interpretability. 
Peptide intensities can instead be aggregated to protein intensities, but this 
is not straightforward due to shared peptides. 

Our package, `bppg` (**b**ipartite **p**eptide **p**rotein **g**raphs), is 
designed to utilise all peptide-level information for protein quantification. 
This is achieved by modelling the relationship between the quantified peptides 
and their proteins of origin using bipartite graphs. This forms the basis of an 
equation system to which an optimisation method is applied in order to best 
explain the measured ratios using the optimised protein ratios. 

The functions and steps of the workflow, from peptide quantities to relative 
protein quantifications, include:

- Reading data of peptide quantities
- Normalization of peptide quantities - reduce technical variability
- Aggregation of replicates per sample group
- Calculation of peptide ratios per group comparison 
- Build bipartite graphs per comparison - preserves peptide-level information
- Optimization step to calculate protein ratios 

In this vignette we included a quick start guide to use `bppg`.

## Background - bppg Theory

### Peptides

There are two categories of peptides. First, **unique** peptides, which 
originate from one protein or protein group (proteins that cannot be 
distinguished based on the measured peptides). Second, **shared** peptides, 
which can derive from more than one protein/protein group. **Unique** peptides
are strong evidence for the presence of a protein and information from
**shared** peptides is more ambiguous. 
`r Citep(bib[["schork2024improvement"]])`

### Bipartite Graphs

Bipartite graphs have two types of nodes (in this case: proteins and peptides),
and the edges exist only between the different types and never within a type
group. A protein node is connected to all their possible 
peptide products. **Unique** peptides therefore have only one edge, while 
**shared** peptides have several. This graph can be based on only *in silico* 
information, creating a theoretical graph.
`r Citep(bib[["schork2024improvement"]])`

This example shows two proteins, which share one peptide and also each have an
unique peptide.
```{r example_plot, echo = FALSE, message = FALSE, out.width="70%", fig.asp=0.9}
file <- system.file("extdata", "quantGraphs_collpept.rds", package = "bppg")
G_coll <- readRDS(file)[[1]][[3]]
plotBipartiteGraph(G_coll)
```


# Installation of `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }

BiocManager::install("bppg")
```


# Asking for help

If you need help with `bppg`, encountered a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r cite}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of `bppg` is protein inference and relative protein quantification
for bottom-up mass spectrometry. This method is only suitable for relative
analysis. Therefore, at least two samples are required to calculate the 
protein quantities. However, at least 3 replicates per group are recommended. 

**Required inputs**:

- Peptide intensities
    - Output from MaxQuant or FragPipe (more options are in development)
    - The read function creates a custom SummarizedExperiment with sample 
    metdata in `colData` and peptide information in `rowData`
- The FASTA file used for peptide identification


## Data

### Example Data
A test dataset and the matching FASTA file are provided within the package. It 
is a subset of a benchmark dataset which was created to determine the 
correctness 
of protein quantification software `r Citep(bib[["ramus_spiked_2016"]])`. 
The original dataset includes: 

- Yeast background, 48 UPS spike-ins
- 9 different spike in concentrations
- 3 replicates each = 27 samples total

The example data used in this vignette includes peptides from 9 yeast proteins.

### Import
The peptide quantification was performed with MaxQuant (version 2.7.3.0)
and resulted in a `peptides.txt` file. The
corresponding function to extract the columns of interest is
`readMqPeptideTable`, this includes the peptide sequences and intensity values.

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides_filtered.txt", package = "bppg")
D <- readMqPeptideTable(path = file)

```

Here is an example of the structure in our SummarizedExperiment dataset:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D  # SummarizedExperiment overview
head(colData(D))  # sample/replicate metadata
assays(D)$intensities[1:5, 1:3]  # peptide intensity matrix 
head(rowData(D))  # peptide metadata (sequence, protein, etc.)
```


## FASTA file

To build the bipartite graph, proteins need to be *in silico* digested and 
peptides matched with their protein of origin. 
For this we have a reduced FASTA file.

```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta",
    package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)
names(fasta)
```

## Build Quantification Graph

Now the theoretical information can be combined with quantification data into
a bipartite peptide protein graph. 
To simplefy this, we created a wrapper function, which includes the digestion 
and the data
preprocessing (normalization, replicate aggreagtion and ratio calculation).
For the generation of the graph protein nodes are collapsed 
(`collProtNodes = TRUE`), this creates protein groups of proteins which are 
indistinguishable by the measured peptides.


```{r build_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- bppg::generateGraphsFromQuantData(D = D,
    fasta = fasta,
    collProtNodes = TRUE)

```

`graphs` is a nested lists, first on the level of comparisons and then the 
multitude of independent graphs for that comparison. The created graph for each
comparison is not fully interconnected but many independet graphs which are
considered separately in the downstream analysis. 

`graphs[[1]][[3]]` shows the first graph of the first comparison 50 amol vs 125
amol. The `igraph` object shows the amount of nodes and edges, the attributes
and the first edges. This example is from RIR2 which does not have an overlap
with another protein. 

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[3]]
```

```{r graphs_plot, message = FALSE, out.width="70%", fig.asp=0.9}
plotBipartiteGraph(graphs[[1]][[3]], node_labels_proteins = "accessions",
    node_labels_peptides = "pep_logRatios")
```


## Optimise Relative Quantification

The optization is performed per graph. The bipartite is the basis for an 
equation system to calculate protein level ratios $R_i$. 
The graph combines information about the peptide ratios ($r_j$) and 
from which proteins they can originate.
However, this is not enough to calculate them direct, instead a configuration
factor ($C_i$) is needed for each protein. 
The optimization algorithm `Rsolnp` (`r Citep(bib[["galanos2025rsolnp"]])`) is 
applied to the resulting equation system.

The optimization is divided into two steps.
First, different protein configurations are tested. 
The best-fitting $C_i$ combination is then extracted in the second step.

Some of the results are not single values but a 
range of equally optimal solutions. This occurs if several combinations return 
the same error and the result is ambiguous. 
Nevertheless, that mirrors the information on the
peptide level, which can be contradicting.

```{r optimization,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
res <- iterateOverCi(graphs[[1]][[1]]) 
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[1]], res)
protQuant
head(assays(protQuant)$results)
```

The result is in a `SummarizedExperiment` and the table can be accesed only via 
the assays function.
In this example the `RiLog_min` and `RiLog_max`, as the `Ci_min` and `Ci_max `
can be seen, which is how range results are encoded. They are the boundaries of 
the log ratio and concentration range for the given error.

## Combine and Export
results from the same comparison can be combiend by using rbind between the 
quant results.

Lastly, the results over all the graphs can be combined into one result. With
each comparison as one assay in the final object.
```{r combine,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
file <- system.file("extdata", "resultsList.rds", package = "bppg")
resultList <- readRDS(file)
SE <- combineComparisons(resultList)
SE
```

and can be exported in an .xlsx file.
```{r export,  echo = TRUE, eval = FALSE, include = TRUE, message = FALSE}
exportSE(SE, file.path("path_to_folder","results.xlsx"))
```

### Parallelization

These examples were quick, but a complete dataset can take hours to go through
all graphs. Since the optimization runs per graph. 
This enables easy parallization, with e.g. 
batchtools `r Citep(bib[["lang2017batchtools"]])`, which supports HPC job 
scheduling. 
An example for this can be seen in our nextflow  workflow for this package:
[here](https://github.com/mpc-bioinformatics/bppg-nf/blob/devel).

# Further information

- information on the detailed workflow, *in silico* analysis and customizing the
graphics are planned for future vignettes.
- more detailed information on the parameters and settings can be found 
in the documentation `r Citep(bib[["schork2026bppg"]])`

# Comparison with other related packages

The amount of R packages for mass spectrometry is continuously increasing. Many
are considering raw spectrum processing or downstream statistical analysis. 
However, the amount of packages that include methods for protein inference and
quantification is limited and mostly confined to simple peptide aggregation 
approaches. 

With `bppg` we provide a more transparent quantification approach that also 
handles proteins without unique peptides.
There are some packages with functionalities for proteomics data that cover 
other parts of the whole workflow. This could include peptide identification 
approaches or further statistical analysis.

- `MSstats` is a package for statistical analysis of proteomics data. Its 
extension `MSstats+` quantifies proteins from peptide abundances by aggregating
them with a weighted sum based on the peak quality.
- `MSnbase` processes proteomics data from raw to quantification data. For the 
protein quantification a median and weighted average function are implemented.
- `msqrob2` analyses differential abundance with an hurdle model. It can be run 
with peptide quantifies, which are later aggregated with `iq::maxLFQ`.
- `mzR` provides parsing for differnt standard file formats used for mass
spectrometry data.
- `DEqMS` is implemented on top of limma and expects protein variance depedend 
on quantified peptides and PSMs.
- `QFeatures` implements the QFeature object consisting of a 
SummarizedExperiment and provides functionality for processing and adding meta 
data.

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```


# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["ole2025biocstyle"]])` with `r CRANpkg("knitr")` 
`r Citep(bib[["xie2025knitr"]])` and `r CRANpkg("rmarkdown")` 
`r Citep(bib[["allaire2026rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` 
`r Citep(bib[["mclean2017refmanager"]])`.


```{r Biblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography

PrintBibliography(bib, .opts = list(hyperlink = "to.bib", style = "html"))
```