---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('bppg')`"
vignette: >
    %\VignetteIndexEntry{1. Introduction to BPPG}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("BiocStyle")
library("RefManageR")
library("bppg")

## Write bibliography information
file <- system.file("extdata", "ref.bib", package = "bppg")
bib <- ReadBib(file, check = FALSE)

bib <- as.BibEntry(c(citation(), citation("BiocStyle")[1], citation("knitr")[1], 
    citation("RefManageR")[1], citation("rmarkdown")[1], 
    citation("sessioninfo")[1], citation("testthat")[1],
    citation("ggplot2")[1], citation("Rsolnp")[1],
    citation("seqinr")[1], citation("batchtools")[1])) + bib

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```



# Introduction

<!-- This is kind of long and should be  moved into the matching section
ALso, add bullet points for steps/what it does and what is good about it -->

Bottom-up mass spectrometry (MS) is an effective tool for analysing
high-throughput proteomics. This method enables the comprehensive investigation
of the proteome of cells, tissues, and organisms. This allows diseases such as 
Parkinson's and the effects of stress on cells and organisms to be studied at a 
proteomic level. 
However, in bottom-up MS, proteins are digested into smaller peptides, which 
are then measured. Consequently, the resulting intensities correspond to 
peptides, rather than proteins. This makes differential analysis on peptide-
level possible, but lacks biological interpretability. Instead peptide 
intensities are aggregated to 
protein intensities, but this is not straight forward due to shared peptides. 

Our package **b**ipartite **p**eptide **p**rotein **g**raphs (`bppg`) aims to
preserve the peptide ratios and built on them for the protein quantification. 
This is done by using bipartite graphs to model the relationship between the 
quantified peptides and their proteins of origin. Thereby building the basis for
an equation system, which an optimization nethods is applied to, to best explain
the measured ratios with protein ratios. This results in calculated protein 
ratios which incorparate shared peptides. 

The functions and steps include:

- Reading data of peptide quantities
- Normalization of peptide quantities - reduce technical variability
- Aggregation of replicates per sample group
- Calculation of peptide ratios per group comparison 
- Build bipartite graphs per comparison - preserves peptide-level information
- Optimization step to calculate protein ratios -
accounts for inconsistencies between peptides

In this vignette we included a quick start guide to use `bppg`.

# Background
## bppg Theory
There are two categories of peptides. First, **unique** peptides, which 
originate from one protein or protein group (proteins that cannot be 
distinguished based on the measured peptides). Second, **shared** peptides, 
which can derive from more than one protein/protein group. **Unique** peptides
are strong evidence for the presence of a protein and information from
**shared** peptides is more ambiguous. 

Bipartite graphs have two types of nodes (in this case: proteins and peptides),
and the edges exist only between the different types and never within a type
group. A protein node is connected to all their possible 
peptide products. **Unique** peptides therefore have only one edge, while 
**shared** peptides have several. This graph can be based on only *in silico* 
information, creating a theoretical graph.
`r Citep(bib[["schork_improvement_2024"]])`

Alternatively, measurements can be 
included, therefore removing all peptides without a valid ratio in the 
sample comparison. This removes proteins without any quantified peptides and
reduces the number of proteins with unique peptide evidence, leading to more
proteins being grouped together.

# Installation of `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
pak::pak("mpc-bioinformatics/bppg")
```


# Asking for help

If you need help with `bppg`, encountered a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r "citation"}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of `bppg` is protein inference and relative protein quantification
for bottom-up mass spectrometry. This method is only suitable for relative
analysis. Therefore, at least two samples are required to calculate the 
protein quantities. However, more replicates per group are recommended. 

**Required inputs**:

- Peptide intensities
    - Output from MaxQuant or FragPipe (more options are in development)
    - The read functionn creates a custom SummarizedExperiment with sample 
    metdata in `colData` and peptide information in `rowData`
- The FASTA file used for peptide identification



## Data

### Example Data
A test dataset and the matching FASTA file are provided with the package. It is
a subset of a benchmark dataset which was created to determine the correctness 
of protein quantification software `r Citep(bib[["ramus_spiked_2016"]])`. 
The original dataset includes: 

- Yeast background, 48 UPS spike-ins
- 9 different spike in concentrations
- 3 replicates each = 27 samples total

The example data used in this vignette includes peptides from 9 yeast proteins.

### Import
The peptide quantification was performed with MaxQuant (version 2.7.3.0)
and resulted in a `peptides.txt` file. The
corresponding function to extract the columns of interest is
`readMqPeptideTable`, this includes the peptide sequences and intensity values.
`remove_contaminants = FALSE` is used to consider contaminants in the downstream
analysis.

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides_filtered.txt", package = "bppg")
D <- readMqPeptideTable(path = file, remove_contaminants = FALSE)

```

The SummarizedExperiment dataset should look as follows:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D  # SummarizedExperiment overview
head(colData(D))  # sample/replicate metadata
assays(D)$intensities[1:5, 1:3]  # peptide intensity matrix 
head(rowData(D))  # peptide metadata (sequence, protein, etc.)
```


## FASTA

To build the bipartite graph, peptides need to be *in silico* digested and 
matched with their protein of origin. For this we have a reduced FASTA file.

```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta",
    package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)
names(fasta)
```

## Build Quantification Graph

Now the theoretical information can be combined with quantification data. 
To simplefy this, we created a wrapper function, which includes the digestion 
and the data
preprocessing (normalization, replicate aggreagtion and ratio calculation).
For the generation of the graph protein nodes are collapsed 
(`collProtNodes = TRUE`), this creates protein groups of proteins which are 
indistinguishable by the measured peptides.


```{r build_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- bppg::generateGraphsFromQuantData(D = D,
    fasta = fasta,
    collProtNodes = TRUE)

```

`graphs` is a nested lists, first on the level of comparisons and then the 
multitude of independent graphs for that comparison. The created graph for each
comparison is not fully interconnected but many independet graphs which are
considered separately in the downstream analysis. 

`graphs[[1]][[1]]` shows the first graph of the first comparison 50 amol vs 125
amol. The `igraph` object shows the amount of nodes and edges, the attributes
and the first edges. This example is from RIR2 which does not have an overlap
with another protein. 

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[1]]
```



## Optimize Relative Quantification

The graph containing peptide ratios is the basis for an equation system to calculate protein 
level ratios. The known variables for this are the peptide ratios ($r_i$) and 
from which proteins they can originate. The unknown variables are the protein 
ratios of the protein quantities of sample A and B ($\frac{B_i}{A_i}=R_i$) and
the proportion of each protein in sample A per connected
graph ($C_i = \frac{A_i}{\sum A_i}$). 
For each peptide, the $R_i$ of the contributing proteins is estimated and used
to reconstruct $r_i$. With $\delta_{ij} = 1$ for contributing proteins and 
$\delta_{ij} = 0$ for not contributing proteins in the graph:

\[ r_j \approx  \sum_{i = 1}^m \left[ 
    \frac{R_i \cdot C_i \cdot \delta_{ij}}{\sum_{k = 1}^m C_k 
        \cdot \delta_{kj}} \right] \]

In general, the 
resulting equation system is underdetermined, since it has more unknown 
variables than known ones. 
Instead an optimization algorithm (`Rsolnp` 
`r Citep(bib[["galanos2025rsolnp"]])`) is 
applied to the system. The error
is defined as the difference between the estimated $R_i$ and reconstructed
$r_i$ and the measured $r_i$.

The optimization is divided into two steps and is performed per graph.

### Optimization
In the first step one $C_i$ is always fixed, while the remaining $C_i$ are 
optimized with regards to $\sum C_i = 1$. 
`iterateOverCi` considers a wide range of possible $C_i$ combination. Besides 
the $C_i$ the protein ratios ($R_i$) are optimized. 
The optimization aims to reduce the error between estimated peptide 
ratios, based on the optimized values, and the measured peptide ratios. With 
the following equation:
\[\underset{R_i, C_i, i=1,\dots,m}{\mathrm{argmin}} 
    \left( \sum_{j=1}^m \varepsilon_j^2  \right)\]

\[ \varepsilon_j = \log(r_j) - \log \left( \sum_{i = 1}^m \left[ 
    \frac{R_i \cdot C_i \cdot \delta_{ij}}{\sum_{k = 1}^m C_k 
        \cdot \delta_{kj}} \right] \right)\]

The result of the first iterations show which `protein` was fixed
to which value (`grid`) and what the log protein ratios (`RLog`) and `error`
were.
The best-fitting $C_i$​ combination is then extracted in the second step.

```{r optimization,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
# graphs[[1]][[1]] has only one proteins and does not need to be optimized
res <- iterateOverCi(graphs[[1]][[2]]) 
head(res)
```

### Extract Best Results

From the previous function the results are filtered for the combination of 
$C_i$s, with the lowest error. Some of the result is not a single value but a 
range of equally optimal solutions. This occurs if several combinations return 
the same error and therefore, the result is ambiguous. 
Nevertheless, that mirrors the information on the
peptide level, which can be contradicting.

```{r results,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[2]], res)
protQuant
head(assays(protQuant)$results)
```

The result is in a `SummarizedExperiment` and the table can be accesed only via 
the assays function.
In this example the `RiLog_min` and `RiLog_max`, as the `Ci_min` and `Ci_max `
can be seen, which is how range results are encoded. They are the boundaries of 
the log ratio and concentration range for the given error.

### Parallelization

The optimization runs per graph. This enables easy parallization, with e.g. 
batchtools `r Citep(bib[["lang2017batchtools"]])`, which supports HPC job scheduling. 
An example for this can be seen in our nextflow  workflow for this package:
[here](https://github.com/mpc-bioinformatics/bppg-nf/blob/devel).

# See Also

Information for *in silico* analysis and plotting can be found in further 
vignettes (currently in preparation)

## Nextflow
We also provide a nextflow workflow 
[here](https://github.com/mpc-bioinformatics/bppg-nf/). Which allows 
for the usage of a wider audience.

# Comparison with other related packages

The amount of R packages for mass spectrometry is continuously increasing. Many
are considering raw spectrum processing or downstream statistical analysis. 
However, the amount of packages that include methods for protein inference and
quantification is limited and mostly confined to simple peptide aggregation 
approaches. 

With `bppg` we provide a more transparent quantification approach that also 
handles proteins without unique peptides.
There are some packages with functionalities for proteomics data that cover 
other parts of the whole workflow. This could include peptide identification 
approaches or further statistical analysis.

- `MSstats` is a package for statistical analysis of proteomics data. Its 
extension `MSstats+` quantifies proteins from peptide abundances by aggregating
them with a weighted sum based on the peak quality.
- `MSnbase` processes proteomics data from raw to quantification data. For the 
protein quantification a median and weighted average function are implemented.
- `msqrob2` analyses differential abundance with an hurdle model. It can be run 
with peptide quantifies, which are later aggregated with `iq::maxLFQ`.
- `mzR` provides parsing for differnt standard file formats used for mass
spectrometry data.
- `DEqMS` is implemented on top of limma and expects protein variance depedend 
on quantified peptides and PSMs.
- `QFeatures` implements the QFeature object consisting of a 
SummarizedExperiment and provides functionality for processing and adding meta 
data.

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```


# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["ole2025biocstyle"]])` with `r CRANpkg("knitr")` 
`r Citep(bib[["xie2025knitr"]])` and `r CRANpkg("rmarkdown")` 
`r Citep(bib[["allaire2026rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` 
`r Citep(bib[["mclean2017refmanager"]])`.


```{r Biblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography

PrintBibliography(bib, .opts = list(hyperlink = "to.bib", style = "html"))
```