---
title: "Introduction to BPPG"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to BPPG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    ggplot2 = citation("ggplot2")[1],
    Rsolnp = citation("Rsolnp")[1],
    seqinr = citation("seqinr")[1]
)

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```

# Introduction

Bottom-up mass spectrometry (MS) is a highly effective tool for the analysis of high
 throughput proteomics. The method facilitates comprehensive investigation of 
 the proteome of cells, tissues and organisms. This enables the study of 
 diseases such as Parkinson's and the effects of stress on a proteomic level.
However, in bottom-up MS the proteins are digested into smaller peptides which
 which are then meassured. Consequently, the resulting abundances are 
 corresponding to peptides. The meassurement of peptides is dependent on the 
 qunatity in the sample, but also physio-chemical characteristics like 
 ionization-efficiancy. The latter means, the measurements of different peptides
 should not be compared directly. Instead differential peptide analysis between
 two sample groups can be performed.


To be able to calculate protein quantities, their peptides have to be aggregated
 in a sensible way. `bppg` aims to do this with the least amount of assumptions
 and simplications. Nevertheless, in many cases it is impossible to distingush
 proteins by the meassured peptides, which are then grouped into a protein group.
 There are two categories of peptides. First, **unique** peptides, which 
 originate from one protein or protein group. Second, **shared** peptides, 
 which can derive from more than one protein/protein group. **Unique** peptides
 are strong evidence for the abundance of a protein and while information from
 **shared** peptides is more ambigious. Bipartite graphs can model the 
 relationship between proteins and peptides, to quantify proteins without 
 **unique** peptides.

Bipartite graphs have two types of nodes (proteins and peptides) and all edges
 are between the two types. A protein node is connected to all their possible 
 peptide products. **Unique** peptides therefore have only one edge, while 
 **shared** peptides have several. This graph can be based on only *insilico* 
 information, creating a theoretical graph. Alternatively, meassurements can be 
 included, therefore removing all peptides without a valid ratio between the 
 two sample groups. This removes proteins without any quantified peptides and
 creates protein groups for missing **shared** peptides.

The quantified graph is the basis for an equation system to calculate protein 
 level ratios. The known factors for this are the peptide ratios (`r_i`) and 
 from which proteins they can originate. The unknwon factors are the protein 
 ratios (`R_i`) and how much each protein contributed to the quantity of their
 peptides compared to the other proteins (`C_i`). `C_i` is only needed for one 
 sample (A), since the other can be calculated with `R_i`. In general, the 
 resulting equation is not solvable, since it has more unknown than knowns. 
 Instead an optimization algorithm (`solnp`) is applied to the system. First,
 different `C_i` are fixed and the remainig `C_i` and `R_i` are optimized for
 the lowest error. The error is calculated on the difference between the 
 calculated `r_i` and the actual ratio.

The optimization allows for inconsistencies between peptides of the same set of
 proteins.

# Installation of  `bppg`

To install the latest version of `bppg` from Bioconductor use the 
following code snippet:

```{r install, eval = FALSE}
pak::pak("mpc-bioinformatics/bppg")
```


# Asking for help

If you need help with `bppg`, encountered a bug or have suggestions for 
improvements leave a post in the 
[Bioconductor support site](https://support.bioconductor.org/) with the 
`bppg` tag or start an issue on our 
[github page](https://github.com/mpc-bioinformatics/bppg).


# Citing `bppg`

We hope that `r Biocpkg("bppg")` will be useful for your research. 
Please use the following information to cite the package and the overall 
approach. Thank you!

```{r "citation"}
## Citation info
citation("bppg")
```

# Quick start to using `bppg`

The main task of bppg is protein inference and relative quantification for 
bottum-up mass spectrometry. This methods is only suitable for relative
analysis.
Therefore, at least two sample groups are requiered to calculate the protein 
quantities. Ideally, with several replicates per sample. 

bppg works with peptide intensities and currently supports the data format from
MaxQuant and McFragger. Otherwise, the dataset can be formatted independtly into
a SummarizedExperiment object with sample/replicate informations in the
columns/colData and peptide information in the rows/rowData.
Additionally, bppg requiers the FASTA file which was used for the peptide
identification.

The second application of bppg is the modeling of peptide and protein 
relationships using bipartite graphs. This is part of the quantification
workflow, but can be used independtly for an insilico analysis.

## Data

### Example Data
A test dataset and the matching FASTA file is provided with the package. 
The data is from ????

- yeast background, UPS spike in
- 9 different spike in concentrations
- 3 replicates each = 27 samples total

### Import
For this dataset the peptide quantificationwas performed with MaxQuant.
Example of loading a `peptides.txt` file:

```{r load_data, echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library("bppg")
library("SummarizedExperiment")

file <- system.file("extdata", "peptides.txt", package = "bppg")
D <- readMqPeptideTable(path = file, LFQ = TRUE, remove_contaminants = FALSE)

```

The SummarizedExperiment dataset should look as followed:

```{r dataset,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D
head(colData(D))
assays(D)$intensities[1:5, 1:3]
head(rowData(D))
```

## Data Preprocessing 

### Normalization

Normalization has to be performed to avoid batch effects. The default method is 
loess, other methods include: median, quantile or lts. 
If the data is already normalized, there is an option for nonorm (no 
normalization). Nevertheless, the function has to be called to allow for 
structural change in the SummarizedExperiment object.

```{r normalize,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
D_norm <- normalizePeptideIntensities(D, method = "loess")


```


### Aggregation
bppg currently only compares the mean of two groups. To avoid calculating a 
large amount of graphs. The replicates need to be assigned their group and then 
there are aggregated, e.g. by mean.
A `missing.limit` can be set to decide for one group and one peptide, if the 
value is regarded as missing, since missing.limit replicates did not quantify 
this peptide. 0.0 allows no missing in the group, while 1.0 allows all missing.

```{r aggregation,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
group <- factor(limma::strsplit2(colnames(D_norm), split = "_")[, 1])
levels(group)
D_aggr <- aggregateReplicates(D_norm, method = "mean", missing.limit = 0.4,
    group = group, seq_col = "Sequence")

```

### Ratio

In the next step the ratios for each group comparison are calculated:

```{r ratio,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
groups  <- levels(group)
peptide_ratios <- calculatePeptideRatios(D = D_aggr, group_levels = groups)

```


## FASTA

To built the bipartite graph peptides need to be insilico digested and matched 
with their protein of origin. This is done by `digestFASTA`, which allows for 
the control of min and max amino acid length of a peptide and how many
miscleavages to consider.
```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(seqinr)
file <- system.file("extdata", "uniprot_test.fasta", package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)

edgelist <- digestFASTA(fasta)

```


## Built Graph
For the graph the quantification data and the FASTA edeglist are combined. 
The `generateQuantGraphs` function creates a comparison specific edgelists, 
including only peptides with valid ratio from that comparison and their 
corresponding proteins.
The option to collapse protein nodes (`collProtNodes`) or collapse peptide nodes
(`collPeptNodes`) is given to reduce the complexity in the graph. Nodes are 
collapsed together if they have the same set of edges. This means collapsing 
protein nodes creates protein groups and therefore, is always recommended. 
`collPeptNodes = TRUE` is  not recommended for quantification purpose, since  
each peptide has its own quantification information. 

```{r built_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- generateQuantGraphs(exp_peptide_ratios = peptide_ratios,
    fasta_edgelist = edgelist, seq_column = "Sequence",
    collProtNodes = TRUE,
    collPeptNodes = FALSE)

```

`graphs` is a list of lists, first on the level of comparisons and then the 
multitude of independtly graphs for that comparison

```{r graphs,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
names(graphs)

graphs[[1]][[1]]
```



## Optimize Realtive Quantification

The optimization is divided into two steps and is performed per graph.

### Optimization
First, different `Ci` percentages are fixed. The `Ci` describes the percentage 
distribution of peptides per graph in sample A.
`iterateOverCi` considers a wide range of possible `Ci` combination by fixing 
one value. The remaining `Ci` are part of the optimization as are the protein 
ratios. The optimization aims to reduce the error between calculated peptide 
ratios, based on the optimized values, and the measured peptide ratios.

```{r optimierung,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
res <- iterateOverCi(graphs[[1]][[1]])
head(res)
```

### Extract Best Results

From the previous results are filtered for the best combination of `Ci`s, with
the lowest error.

```{r results,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
protQuant <- automatedAnalysisIteratedCi(graphs[[1]][[1]], res)
head(assays(protQuant)$results)
```


### Parallization

The optimization runs per graph. This enables easy parallization, for with
example `batchtools`. an example for this can be seen in our nextflow workflow 
for this package.

# Further information

- information for insilico analysis and plotting can be found in further 
vignettes

## Nextflow
We also provide a nextflow workflow here. which allows for usage of a wider 
audiance

# Comparison with other related packages
what is this research?

# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

