---
title: "Analysing *in silico* graphs"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
  - name: Julika Wenzel
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: julika.wenzel@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{2. Analysing *in silico* graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## Related to 
    ## https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    ggplot2 = citation("ggplot2")[1],
    seqinr = citation("seqinr")[1]
)

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")

```

# Introduction
Besides the general workflow bppg offers the option to calculate the 
theoretical graph, if all peptides are measured.
This allows the possibility to analyse characteristics of the *in silico* 
graphs and to compare it with the data depedend graph.

# Process FASTA
As the first step we need the proteom aka FASTA file for an organisms of
intereset.
In this example we are using a subset of *S. cerevisiae*, to illustrate 
some function on a small dataset.

The processing follows the general workflow by *in silico* digesting the
proteins in the FASTA file. Which returns an edgelist with pairs of peptides
and proteins.

The protOrigin can be added to 

```{r digest,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
library(bppg)
library(seqinr)

file <- system.file("extdata", "uniprot_proteome_Scerevisiae_filtered.fasta",
    package = "bppg")
fasta <- seqinr::read.fasta(file = file, seqtype = "AA", as.string = TRUE)

origin <- as.list(rep("yeast", 9))
edgelist <- digestFASTA(fasta, protOrigin = origin)
head(edgelist)
```


# Built graph
The edgelist without quantification values can be formed into bipartite
graphs with `generateGraphsFromEdgelist`. This function only needs the 
protein - peptide edgelist and gives the option to collapse protein and/or
peptide nodes. 
We recommend to always use `collProtNodes = TRUE`, since it groups proteins
that can not be distinguished based on their peptides. In this case it 
does not matter, since all peptides are used and each protein should have
at least one unique peptide. Otherwise their would be identical sequence 
entries.

For `collPeptNodes` both options (`TRUE`, `FALSE`) can be informative.
`collPeptNodes = FALSE` gives the option to analyse how many **unique** and
**shared** a protein can have in theory. It gives insights how many peptides
overlap with other proteins and if an unique identifier is likely. 
With `collPeptNodes = TRUE` questions about structure and solvability can be
anwsered. Which is what we will do in this tutorial.

```{r build_graph,  echo = TRUE, eval = TRUE, include = TRUE, message = FALSE}
graphs <- generateGraphsFromEdgelist(
    edgelist = edgelist,
    collProtNodes = TRUE,
    collPeptNodes = TRUE)

graphs
```

# Analyse graph
While this example is a simplefied version with only four graphs, the analysis 
can be highlighted. On the one hand, the topolgy of the graphs can be studied.
On the other hand, the relationship and statistics of the proteins and peptides
can be analysed. 

The general topolgy and isomorphisms of the graphs can be analyzed with 
`.generatePrototyeList`. This function identifies graphs of the same structure
and creates a list of prototype graphs and how often their structure occurs in 
the given input. These can also be plotted as described in 
vingette `Visualize bppg`.

```{r isomorphism}

```

Secondly, we can look at graph statistics per graph and over all graphs. How
many proteins and peptides are there, how many edges. However, also regarding
**shared** and **unique** peptides and how many protein do not have an unique
peptide.

```{r graphStats}

```



# R session information

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

