---
title: "Visualisation with BPPG"
author: 
  - name: Leonie Laarmann
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: leonie.laarmann@rub.de

output: bipartite graphs as pngs
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{Visualisation with BPPG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

title: "Visualisation with BPPG"
output: BiocStyle::html_document
vignette: >
%\VignetteIndexEntry{Visualisation with BPPG}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
-------------------------

## Bib setup

```{r}
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    ggplot2 = citation("ggplot2")[1],
    igraph = citation("igraph")[1],
    ggraph = citation("ggraph")[1]
)

BibOptions(max.names = 2, bib.style = "authoryear", style = "citation")
```

# Introduction

This vignette provides an introduction to visualizing bipartite graphs using the BPPG package.

# Visualizing Bipartite Graphs

The BPPG package offers a range of functions for visualizing bipartite peptide-protein graphs.
In this section, we will explore some of the key visualization options and how to apply them to your data.

You may use the `.plotBipartiteGraph()` function to create a basic visualization of your bipartite graph.
The function takes a bipartite `igraph` graph object as input and generates a plot that represents the relationships between peptides and proteins.

There are several customization options available, such as adjusting node labels, node shapes, colors, and the legend.

## Customization

# Node Labels

The labels of the nodes can be customized using the `node_labels_proteins` and `node_labels_peptides` parameters.
For proteins, the available options are `"letters"`, `"accessions"` and `"numbers_noord"`.
For peptides, the available options are `"numbers"`, `"sequences"`, `"pep_ratios"` and `"pep_ratio_aggr"`.

For example:

```{r}
plotBipartiteGraph(
    G,
    node_labels_proteins = "accessions",
    node_labels_peptides = "sequences"
)
```

When peptide ratios are used as labels, the number of displayed decimal places can be specified using `round_digits`.

# Node Shapes

The shape of the nodes can be customized using the `three_shapes` parameter.
By default, proteins and peptides are displayed using different shapes. Setting `three_shapes` to `TRUE` additionally distinguishes between shared and unique peptides.

```{r}
plotBipartiteGraph(
    G,
    three_shapes = TRUE
)
```

When 'three_shapes' = FALSE, proteins and peptides are displayed using different shapes, but shared and unique peptides are not distinguished. 

# Colors and Edge Width

The colors of the nodes can be customized using the `vertex.color` parameter.
The colors are specified for proteins, shared peptides, and unique peptides, respectively.
The default colors are specifically chosen to be colorblind-friendly.

The width of the edges can be adjusted using the `edge.width` parameter.

```{r}
plotBipartiteGraph(
    G,
    vertex.color = c("red", "blue", "green"),
    edge.width = 2
)
```

# Legend

The legend may be hidden using the `legend` parameter.
By default, the legend is displayed, but it can be hidden by setting this parameter to `FALSE`.

```{r}
plotBipartiteGraph(
    G,
    legend = FALSE
)
```

# Saving the Plot

The generated plot can be saved as a JPEG file by providing a path to `save_path`.

```{r}
plotBipartiteGraph(
    G,
    save_path = "bipartite_graph.jpg"
)
```

# Summary

Plotting bipartite graphs with the BPPG package provides a way to visualize relationships between peptides and proteins.
By customizing node labels, node shapes, colors, edge widths, and legends, informative visualizations of the graph structure can be created.
